Estimated fitness effects of amino-acid mutations to SARS-CoV-2 proteins

### 2023-01-12 paper/paper.tex
draft of paper figures

>>>>

\subsection{Theory}
\jdbcomment{Include Richard's calculations, either in entirety or summarized with details in Methods / Appendix}

\begin{figure*}
{\bf \Large A \hspace{0.47\linewidth} B}

\includegraphics[width=0.48\linewidth,valign=t]{figs/schematic/schematic.pdf}
\hspace{0.02\linewidth}
\includegraphics[width=0.5\linewidth,valign=t]{figs/avg_counts.pdf}
\caption{
Calculation of expected versus actual counts for each mutation.
{\bf (A)}
The number of expected counts of each type of nucleotide mutation is computed from the observed counts at four-fold degenerate sites,
and then compared the actual observed counts of each mutation.
Note that the counts represent independent occurrences of each mutation on the tree, not counts of mutations in the alignment of observed sequences (a single occurrence of a mutation can be present in many sequences in an alignment).
{\bf (B)}
Expected versus actual counts for each nucleotide mutation type averaged across all sites where the mutation is four-fold degenerate, synonymous (including four-fold degenerate), nonsynonymous, or introduces a stop codon.
Synonymous mutations are often roughly neutral, so the number of expected and actual counts are similar, but nonsynonymous and especially stop-codon mutations are often deleterious, so the number of actual counts is generally less than the number of expected counts.
Note that due to the highly uneven mutation spectrum of SARS-CoV-2, some nucleotide mutation types are expected to occur far more often than others.
An interactive version of panel B is at \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html}.
\label{fig:expected_vs_actual}
}
\end{figure*}

\begin{figure*}
{\bf \Large A}

\includegraphics[width=\linewidth, trim={0in 0.7in 0 0.05in}, clip=true]{figs/effects_histogram.pdf}

{\bf \Large B}

\includegraphics[width=\linewidth, trim={0in 0.36in 0 0in}, clip=true]{figs/effects_dist.pdf}
\caption{
Distribution of effects of different types of mutations.
{\bf (A)}
Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
Neutral mutations have effects of zero, and deleterious mutations have negative effects.
\jdbcomment{Maybe add dotted vertical line at zero to denote neutrality?}
{\bf (B)}
Distribution of effects of each type of mutation for each viral gene.
The dark squares indicate the median effect of mutations of that type for that gene, and the lighter rectangles span the interquartile range.
Mutation types are color-coded as in panel A.
Both panels show only mutations with expected counts of at least 20.
For interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene), see \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html}.
\label{fig:effects_dist}
}
\end{figure*}

\begin{figure*}
\centering
{\bf \Large A \hspace{0.32\linewidth} B \hspace{0.32\linewidth}}

\includegraphics[width=0.32\linewidth]{figs/clade_corr_chart.pdf}
\hspace{0.05\linewidth}
\includegraphics[width=0.32\linewidth]{figs/subset_corr_chart.pdf}
\caption{
Correlations between amino-acid fitness estimates made using different sequence subsets.
{\bf (A)} Estimates made using only sequences from the Omicron BA.2 clade or only the Alpha clade, or {\bf (B)} estimates made using only sequences from the USA or England.
In the scatter plots, each point is a different amino-acid mutation, and the orange text at upper left shows the number of mutations and Pearson correlation coefficient.
Both panels show only mutations with expected counts of at least 20.
For interactive versions of these plots that show correlations among estimates from all clade pairs, allow adjustment of the expected-count cutoff, and allow examination of correlations for just specific genes, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html}.
\label{fig:corr}
}
\end{figure*}

\begin{figure*}
{\bf \Large A}

\includegraphics[width=0.9\linewidth]{figs/dms_S_corr.pdf}
\vspace{0.05in}

{\bf \Large B}

\includegraphics[width=0.9\linewidth]{figs/dms_nsp5_corr.pdf}
\caption{
Correlation of fitness effect estimates with experimental measurements from deep mutational scanning studies for {\bf (A)} spike and {\bf (B)} Mpro (nsp5).
Each point is an amino-acid mutation, and the oragne text in the upper left shows the number of mutations and Pearson correlation coefficient.
Data are taken from two independent deep mutational scanning studies for spike~\citep{cite}, and two studies for Mpro~\citep{cite}.
Each sub-panel shows a different set of mutations (those that are measured in each of the two datasets being correlated), and uses a minimum expected count cutoff of 20.
Interactive versions of these plots that enable subsetting on just mutations shown across all datasets and adjustment of the minimum expected count cutoff are at \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html}.
\label{fig:dms_corr}
}
\end{figure*}

\begin{figure*}
{\bf \Large A}

E protein mutation effects with minimum expected counts of 20 and \textbf{x} marking wildtype identity in Wuhan-Hu-1.

\includegraphics[width=\linewidth, trim={0 0 0 0.4in}, clip=true]{figs/E_view1.pdf}

{\bf \Large B}

E protein mutation effects with minimum expected counts of 5 and \textbf{x} marking wildtype identity in BQ.1.

\includegraphics[width=\linewidth, trim={0 0 0 0.4in}, clip=true]{figs/E_view2.pdf}
\caption{
Effects of amino-acid mutations to E protein.
{\bf (A)} and {\bf (B)} provide two views of the estimated mutation effects with different thresholds for the minimum expected counts and different references used to mark the ``wildtype'' identity, as indicated in the titles above the plots.
For both panels, the area plot at top shows the average effects of mutations at each site, and the heatmap shows the effects of specific amino acids.
An interactive version of this plot at \url{https://jbloomlab.github.io/SARS2-mut-fitness/E.html} enables zooming, mouseovers, and exploration of the data.
Comparable interactive plots for all other viral proteins are linked at \jdbcomment{create landing pad page with these links}.
\label{fig:E}
}
\end{figure*}

\begin{figure*}
{\bf \Large A}

\includegraphics[width=0.8\linewidth]{figs/clade_fixed_muts_hist.pdf}
\vspace{0.05in}

{\bf \Large B}

\includegraphics[width=\linewidth]{figs/clade_fixed_muts.pdf}

\caption{
Effects of amino-acid mutations that have fixed in at least one SARS-CoV-2 clade.
{\bf (A)}
Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1 and all mutations that are fixed in at least one of the clades (as defined by Nextclade\jdbcomment{cite}).
\jdbcomment{add vertical line at zero to these plots.}
{\bf (B)}
Effects of individual fixed mutations faceted by whether they are in spike or another protein.
Plots show estimates of effects of mutations made across all clades (including those that have fixed the mutation) and just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
Estimates are only shown for mutations with expected counts of at least 20.
Interactive versions of this plot that enable easier exploration of the data are at \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts.html}.
\label{fig:fixed}
}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.8\linewidth, trim={0 1.1in 0 0in}, clip=true]{figs/fitness_vs_terminal.pdf}
\caption{
Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
Each point is an amino-acid mutation, and the orange text in the upper right give the number of mutations and the Pearson correlation coefficient.
This plot shows only mutations with at least 20 expected counts and 5 actual counts.
An interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene, is at \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html}.
\jdbcomment{add same color coding as in other panels for synonymous, nonsynonymous, stop types}
\label{fig:terminal}
}
\end{figure*}
